Diversity dynamics in Nymphalidae butterflies: Effect of phylogenetic uncertainty on diversification rate shift estimates

### 2015-01-14 MS_revised.tex
revision 2

Hostplant shifts have been invoked to be responsible for a great part of
the biodiversity of herbivorous insects \cite{mitter1988}. The study of the
evolution of hostplant use has spawned several theories explaining the
evolutionary interactions between plants and insects \cite{nyman2012}. The
``escape-and-radiate'' hypothesis \cite{ehrlich1964}, the ``oscillation
hypothesis'' \cite{janz2011, nylin2014} or ``diffuse cospeciation'' \cite{nyman2012} and the
``musical chairs hypothesis'' \cite{hardy2014}.

@@

If there is strong phylogenetic signal for increases or decreases in net
diversification rates for a node, it is expected that these shifts would
be inferred by MEDUSA in most of the posterior distribution of trees.
However, weak phylogenetic signal for some nodes can cause some clades
to be absent in some trees and MEDUSA will be unable to estimate any
diversification shift (due to a non-existent node).
The relatively low phylogenetic support for many nodes in the Nymphalidae
tree is likelyl the reason
why MEDUSA estimated net diversification rate shifts with a probability
higher than 0.90 in the sample of trees for only four shifts: the genus
\emph{Charaxes}, the genus \emph{Ypthima}, part of Ithomiini and the
clade \emph{Callicore} + \emph{Diaethria} (Fig. 3), while estimating
shifts for other lineages with much lower probability.

@@

\subsubsection*{Danaini}
Our MultiMEDUSA approach showed a significant slowdown in net
diversification rate in the subtribe Danaina of the Danini. Both Danaina
and the sister clade Euploeina feed mainly on Apocynaceae and thus a
hostplant shift should not be responsible for the observed slowdown of
diversification in the Danaina. As expected, our BiSSE analysis of
Apocynaceae feeders shows that there is no effect of feeding on this
plant family on the net diversification rates of Nymphalidae lineages.
Many of the Danaina are large, strong fliers, highly migratory and
involved in mimicry rings, including the best-known migratory butterfly,
the monarch (\emph{Danaus plexippus}).
The causes for a lower net diversification rate in the
Danaina remains to be investigated, but their great dispersal power
might be involved in preventing allopatric speciation. It has been found
in highly vagile species in the nymphalid genus \emph{Vanessa} that
dispersal has homogenized populations due to gene flow, as 
vagile species seem to be genetically homogeneous in their
populations \cite{wahlberg2011}.

@@

Our MultiMEDUSA approach to perform analyses on the posterior
distribution of trees found strong support for an increase in net
diversification rate in the tribe Ithomiini, the genera \emph{Ypthima},
\emph{Charaxes}, the clade \emph{Callicore} + \emph{Diaethria} and for a
decrease in net diversification rate in the
subtribe Danaina. Due to phylogenetic uncertainty, we did not obtain
strong support for other diversification shifts in Nymphalidae. Our
BiSSE analysis corroborated other studies in that Solanaceae-feeding,
or a codistributed character, was likely important in the
diversification of Ithomiini butterflies. However, by applying MEDUSA we
found that a critical character in the radiation of the group might have
appeared after the shift from Apocynaceae to Solanaceae. We also found
that Apocynaceae-feeding is not responsible for the
slowdown of diversification in Danaina. Ithomiini and Solanaceae
diversified near simultaneously, which is in agreement with the diffuse
cospeciation hypothesis \cite{nyman2012, janz2011}.

>>>>

Hostplant shifts have been invoked to be responsible for a great part of
the biodiversity of herbivorous insects \cite{mitter1988}. The study of the
evolution of hostplant use has spawned several theories explaining the
evolutionary interactions between plants and insects \cite{nyman2012}: the
``escape-and-radiate'' hypothesis \cite{ehrlich1964}, the ``oscillation
hypothesis'' \cite{janz2011, nylin2014} or ``diffuse cospeciation'' \cite{nyman2012} and the
``musical chairs hypothesis'' \cite{hardy2014}.

@@

If there is strong phylogenetic signal for increases or decreases in net
diversification rates for a node, it is expected that these shifts would
be inferred by MEDUSA in most of the posterior distribution of trees.
However, weak phylogenetic signal for some nodes can cause some clades
to be absent in some trees and MEDUSA will be unable to estimate any
diversification shift (due to a non-existent node).
The relatively low phylogenetic support for many nodes in the Nymphalidae
tree is likely the reason
why MEDUSA estimated net diversification rate shifts with a probability
higher than 0.90 in the sample of trees for only four shifts: the genus
\emph{Charaxes}, the genus \emph{Ypthima}, part of Ithomiini and the
clade \emph{Callicore} + \emph{Diaethria} (Fig. 3), while estimating
shifts for other lineages with much lower probability.

@@

\subsubsection*{Danaini}
Our MultiMEDUSA approach showed a significant slowdown in net
diversification rate in the subtribe Danaina of the Danini. Both Danaina
and the sister clade Euploeina feed mainly on Apocynaceae and thus a
hostplant shift should not be responsible for the observed slowdown of
diversification in the Danaina. As expected, our BiSSE analysis of
Apocynaceae feeders shows that there is no effect of feeding on this
plant family on the net diversification rates of Nymphalidae lineages.
Many of the Danaina are large, strong fliers, highly migratory and
involved in mimicry rings, including the best-known migratory butterfly,
the monarch (\emph{Danaus plexippus}).
The causes for a lower net diversification rate in the
Danaina remains to be investigated, but their great dispersal power
might be involved in preventing allopatric speciation. It has been found
in highly vagile species in the nymphalid genus \emph{Vanessa} that
dispersal has homogenized populations due to gene flow, as 
vagile species seem to be genetically homogeneous among
populations \cite{wahlberg2011}.

@@

Our MultiMEDUSA approach to perform analyses on the posterior
distribution of trees found strong support for an increase in net
diversification rate in the tribe Ithomiini, the genus \emph{Ypthima},
genus \emph{Charaxes}, the clade \emph{Callicore} + \emph{Diaethria} and for a
decrease in net diversification rate in the
subtribe Danaina. Due to phylogenetic uncertainty, we did not obtain
strong support for other diversification shifts in Nymphalidae. Our
BiSSE analysis corroborated other studies in that Solanaceae-feeding,
or a codistributed character, was likely important in the
diversification of Ithomiini butterflies. However, by applying MEDUSA we
found that a critical character in the radiation of the group might have
appeared after the shift from Apocynaceae to Solanaceae. We also found
that Apocynaceae-feeding is not responsible for the
slowdown of diversification in Danaina. Ithomiini and Solanaceae
diversified near simultaneously, which is in agreement with the diffuse
cospeciation hypothesis \cite{nyman2012, janz2011}.
